---
title: "21 Choropleth Mapping with R"
format: html
author: "Trinh Minh Tam"
date: "Jun 16, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
editor: 
  markdown: 
    wrap: 72
---

## 21.1 Overview

Choropleth mapping involves the symbolisation of enumeration units —
such as countries, provinces, states, counties, or census units — using
area patterns or graduated colours. For example, a social scientist may
use a choropleth map to portray the spatial distribution of the aged
population of Singapore by Master Plan 2014 Subzone Boundary.

In this chapter, you will learn how to plot functional and truthful
choropleth maps using an R package called
[**tmap**](https://r-tmap.github.io/tmap/).

::: callout-tip
It is advisable to read the functional description of each function
before using it.
:::

## 21.2 Getting Started

In this hands-on exercise, the key R package used is
[**tmap**](https://cran.r-project.org/web/packages/tmap/). Besides
**tmap**, four other R packages will be used:

- [**readr**](https://readr.tidyverse.org/) for importing delimited text
  files,
- [**tidyr**](https://tidyr.tidyverse.org/) for tidying data,
- [**dplyr**](https://dplyr.tidyverse.org/) for wrangling data, and
- [**sf**](https://cran.r-project.org/web/packages/sf/index.html) for
  handling geospatial data.

Among the four, **readr**, **tidyr**, and **dplyr** are part of the
**tidyverse** package.

The code chunk below will install and load these packages in RStudio.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

::: callout-note
Notice that we only need to install **tidyverse** rather than **readr**,
**tidyr**, and **dplyr** individually.
:::

## 21.3 Importing Data into R

### 21.3.1 The Data

Two datasets will be used to create the choropleth map:

- **Master Plan 2014 Subzone Boundary (Web)** (i.e.
  `MP14_SUBZONE_WEB_PL`) in ESRI shapefile format. It can be downloaded
  from [data.gov.sg](https://data.gov.sg/). This is a geospatial dataset
  consisting of the geographical boundary of Singapore at the planning
  subzone level. The data is based on the URA Master Plan 2014.

- **Singapore Residents by Planning Area / Subzone, Age Group, Sex and
  Type of Dwelling, June 2011-2020** in CSV format (i.e.
  `respopagesextod2011to2020.csv`). This is an aspatial data file,
  downloadable from the [Department of Statistics,
  Singapore](https://www.singstat.gov.sg/). Although it does not contain
  any coordinate values, its `PA` and `SZ` fields can be used as unique
  identifiers to geocode to the `MP14_SUBZONE_WEB_PL` shapefile.

### 21.3.2 Importing Geospatial Data into R

The code chunk below uses the `st_read()` function of the **sf** package
to import the `MP14_SUBZONE_WEB_PL` shapefile into R as a simple feature
data frame called `mpsz`.

```{r}
mpsz <- st_read(dsn = "../Data/geospatial",
                layer = "MP14_SUBZONE_WEB_PL")
```

You can examine the content of `mpsz` using the code chunk below.

```{r}
mpsz
```

::: callout-note
Notice that only the first ten records will be displayed. This is
because that's the default max.print setting for sf objects — it's a
readability safeguard, not a limit on the data itself. The underlying
object still contains all 323 features.
:::

### 21.3.3 Importing Attribute Data into R

Next, we will import the `respopagesextod2011to2020.csv` file into
RStudio and save it as an R data frame called `popdata`.

This task will be performed using the `read_csv()` function of the
**readr** package, as shown below.

```{r}
popdata <- read_csv("../Data/respopagesextod2011to2020.csv")
```

### 21.3.4 Data Preparation

Before a thematic map can be prepared, you are required to build a data
table containing year-2020 values. The data table should include the
variables `PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, and
`DEPENDENCY`:

- **YOUNG**: age group 0-4 up to age group 20-24,
- **ECONOMY ACTIVE**: age group 25-29 up to age group 60-64,
- **AGED**: age group 65 and above,
- **TOTAL**: all age groups, and
- **DEPENDENCY**: the ratio of the young and aged populations against
  the economically active group.

#### 21.3.4.1 Data Wrangling

The following data-wrangling and transformation functions will be used:

- `pivot_wider()` of the **tidyr** package, and
- `mutate()`, `filter()`, `group_by()`, and `select()` of the **dplyr**
  package.

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from = AG,
              values_from = POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         + rowSums(.[12])) %>%
  mutate(`ECONOMY ACTIVE` = rowSums(.[7:11]) +
           rowSums(.[13:15])) %>%
  mutate(`AGED` = rowSums(.[16:21])) %>%
  mutate(`TOTAL` = rowSums(.[3:21])) %>%
  mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
         / `ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`,
         `ECONOMY ACTIVE`, `AGED`,
         `TOTAL`, `DEPENDENCY`)
```

#### 21.3.4.2 Joining the Attribute Data and Geospatial Data

Before we can perform the geo-relational join, one extra step is
required: convert the values in the `PA` and `SZ` fields to uppercase.
This is because the values of `PA` and `SZ` contain a mix of upper- and
lowercase letters, whereas `SUBZONE_N` and `PLN_AREA_N` are in
uppercase.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, `left_join()` of **dplyr** is used to join the geographical data
and attribute table using the planning subzone name (i.e. `SUBZONE_N`
and `SZ`) as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

::: callout-note
#### Things to learn from the code chunk above:

- `left_join()` of **dplyr** is used with the `mpsz` simple feature data
  frame as the left table to ensure that the output is also a simple
  feature data frame.
:::

```{r}
write_rds(mpsz_pop2020, "../Data/mpszpop2020.rds")
```

## 21.4 Choropleth Mapping Geospatial Data Using *tmap*

Two approaches can be used to prepare a thematic map with *tmap*:

- Plotting a thematic map quickly using `qtm()`.
- Plotting a highly customisable thematic map using **tmap** elements.

### 21.4.1 Plotting a Choropleth Map Quickly Using `qtm()`

The easiest and quickest way to draw a choropleth map with **tmap** is
by using `qtm()`. It is concise and provides good default visualisation
in many cases.

The code chunk below draws a cartographically standard choropleth map.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```

::: callout-note
### Things to learn from the code chunk above:

- `tmap_mode()` with the `"plot"` option produces a static map. For
  interactive mode, use the `"view"` option.
- The `fill` argument is used to map the attribute (i.e. `DEPENDENCY`).
:::

### 21.4.2 Creating a Choropleth Map Using *tmap*'s Elements

Despite its usefulness for drawing a choropleth map quickly and easily,
the disadvantage of `qtm()` is that it makes the aesthetics of
individual layers harder to control. To draw a high-quality cartographic
choropleth map, **tmap**'s drawing elements should be used.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by Planning Subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics (DOS)",
             position = c("left", "bottom"))
```

In the following sub-sections, we will share with you the **tmap**
functions used to plot these elements.

#### 21.4.2.1 Drawing a Base Map

The basic building block of **tmap** is `tm_shape()` followed by one or
more layer elements such as `tm_fill()` and `tm_polygons()`.

In the code chunk below, `tm_shape()` defines the input data (i.e.
`mpsz_pop2020`) and `tm_polygons()` draws the planning subzone polygons.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

#### 21.4.2.2 Drawing a Choropleth Map Using `tm_polygons()`

To draw a choropleth map showing the geographical distribution of a
selected variable by planning subzone, we simply assign the target
variable (e.g. `Dependency`) to `tm_polygons()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY")
```

::: callout-note
#### Things to learn from `tm_polygons()`:

- The default interval binning used to draw the choropleth map is called
  `"pretty"`. A detailed discussion of the data-classification methods
  supported by **tmap** is given in the next sub-section.
- The default colour scheme is `YlOrRd` from ColorBrewer. You will learn
  more about colour schemes shortly.
- By default, missing values are shaded in grey.
:::

#### 21.4.2.3 Drawing a Choropleth Map Using `tm_fill()` and `tm_borders()`

In fact, `tm_polygons()` is a wrapper around `tm_fill()` and
`tm_borders()`. `tm_fill()` shades the polygons using the default colour
scheme, and `tm_borders()` adds the borders of the shapefile onto the
choropleth map.

The code chunk below draws a choropleth map using `tm_fill()` alone.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY")
```

Notice that the planning subzones are shaded according to their
respective dependency values.

To add the boundaries of the planning subzones, `tm_borders()` is used,
as shown below.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,
             fill_alpha = 0.1)
```

Notice that light-grey border lines have been added to the choropleth
map.

The `alpha` argument defines transparency as a number between 0 (fully
transparent) and 1 (fully opaque). By default, the alpha value of `col`
is used (normally 1).

Besides `alpha`, there are three other arguments for `tm_borders()`:

- `col` — border colour,
- `lwd` — border line width (default is 1), and
- `lty` — border line type (default is `"solid"`).

### 21.4.3 Data Classification Methods of **tmap**

Most choropleth maps employ some method of data classification. The
point of classification is to take a large number of observations and
group them into data ranges or classes.

**tmap** provides ten data-classification methods: `fixed`, `sd`,
`equal`, `pretty` (default), `quantile`, `kmeans`, `hclust`, `bclust`,
`fisher`, and `jenks`.

To specify a data-classification method, the `style` argument of
`tm_fill()` or `tm_polygons()` is used.

#### 21.4.3.1 Plotting Choropleth Maps with Built-in Classification Methods

The code chunk below shows a Jenks data classification using 5 classes.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

The next code chunk uses the `equal` classification method.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

Notice that the distribution produced by the quantile method is more
evenly distributed than the one produced by the equal method.

::: callout-warning
**Maps Lie!**

**DIY:** Using what you have learned, prepare choropleth maps using
different classification methods supported by **tmap** and compare the
differences.

**DIY:** Prepare choropleth maps using the same classification method
but with different numbers of classes (e.g. 2, 6, 10, 20). Compare the
resulting maps — what can you observe?
:::

#### DIY 1: Comparing Different Classification Methods

The code chunk below produces four choropleth maps of the `DEPENDENCY`
variable using four different classification styles, all with `n = 5`
classes.

```{r}
#| fig-width: 12
#| fig-height: 10

equal_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Equal Interval")

quantile_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Quantile")

jenks_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "jenks",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Jenks (Natural Breaks)")

sd_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "sd",
                n = 5,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Standard Deviation")

tmap_arrange(equal_map, quantile_map, jenks_map, sd_map,
             asp = 1, ncol = 2)
```

**Observations:**

- **Equal Interval** divides the data range into bins of equal width.
  Because the `DEPENDENCY` variable contains a few extreme outliers (max
  ≈ 19), most subzones end up squeezed into the lowest class, leaving
  the map looking flat and uninformative.
- **Quantile** places an equal number of subzones in each class. This
  produces a visually balanced map with strong colour contrast, but it
  can exaggerate small differences between similar values.
- **Jenks (Natural Breaks)** finds break points that minimise
  within-class variance and maximise between-class variance. The
  resulting map highlights genuine clusters in the data and is often the
  best general-purpose choice.
- **Standard Deviation** centres the classes around the mean. It is
  useful for revealing how unusual each subzone is relative to the
  average, but it relies on the data being roughly normally distributed.

The key lesson: the same data can produce very different-looking maps
depending on the classification method. A cartographer's choice can
mislead the reader — hence "Maps Lie".

#### DIY 2: Varying the Number of Classes

The code chunk below produces four quantile choropleth maps with `n = 2, 6, 10, and 20` classes.

```{r}
#| fig-width: 12
#| fig-height: 10

n2_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 2,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 2 classes")

n6_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 6,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 6 classes")

n10_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 10,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 10 classes")

n20_map <- tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 20,
                values = "brewer.blues")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("n = 20 classes")

tmap_arrange(n2_map, n6_map, n10_map, n20_map,
             asp = 1, ncol = 2)
```

**Observations:**

- **n = 2** produces a binary map — every subzone is either "above" or "below" the median. Spatial patterns are easy to see at a glance, but most of the detail is lost.
- **n = 6** strikes a reasonable balance between readability and detail. Broad regional patterns remain visible while finer variation is still distinguishable.
- **n = 10** begins to reveal more local variation, but the human eye struggles to differentiate so many similar shades of blue, especially on small polygons.
- **n = 20** is essentially unreadable as a thematic map. The colour ramp is too granular for visual comparison, and adjacent classes become indistinguishable.

**Rule of thumb:** most cartographers recommend **between 4 and 7 classes** for a sequential colour scheme. Fewer classes oversimplify the data; more classes overwhelm the reader.

#### 21.4.3.2 Plotting a Choropleth Map with Custom Breaks

For all the built-in styles, the category breaks are computed
internally. To override these defaults, the breakpoints can be set
explicitly via the `breaks` argument of `tm_fill()`. Note that in
**tmap**, the breaks include both a minimum and a maximum. As a result,
to end up with *n* categories, *n + 1* elements must be specified in the
`breaks` vector (the values must be in increasing order).

Before we begin, it is good practice to compute descriptive statistics
for the variable before setting breakpoints. The code chunk below
computes and displays the descriptive statistics of the `DEPENDENCY`
field.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

Based on the results, we set breakpoints at 0.60, 0.70, 0.80, and 0.90.
In addition, we include a minimum and maximum, which we set at 0 and
1.00. Our `breaks` vector is therefore
`c(0, 0.60, 0.70, 0.80, 0.90, 1.00)`.

Now we plot the choropleth map using the code chunk below.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
              breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

### 21.4.4 Colour Scheme

**tmap** supports colour ramps that are either user-defined or chosen
from a set of predefined colour ramps from the **RColorBrewer** package.

#### 21.4.4.1 Using ColorBrewer Palettes

To change the colour, assign the preferred colour to the `values`
argument of `tm_scale_intervals()`, as shown below.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

Notice that the choropleth map is now shaded in green.

To reverse the colour shading, add a `"-"` prefix.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

### 21.4.5 Map Layouts

Map layout refers to the combination of all map elements into a cohesive
map. Map elements include, among others, the objects to be mapped, the
title, the scale bar, the compass, margins, and aspect ratios. Colour
settings and data-classification methods (covered earlier) relate to the
palette and breakpoints, which affect how the map looks.

#### 21.4.5.1 Map Legend

In **tmap**, several `tm_legend()` options are provided to change the
placement, format, and appearance of the legend.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by Planning Subzone \n(Jenks classification)")
```

#### 21.4.5.2 Map Style

**tmap** allows a wide variety of layout settings to be changed via
`tmap_style()`.

The code chunk below uses the `classic` style.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

#### 21.4.5.3 Cartographic Furniture

Besides map style, **tmap** also provides arguments for drawing other
map furniture such as compass, scale bar, and grid lines.

In the code chunk below, `tm_compass()`, `tm_scale_bar()`, and
`tm_grid()` are used to add a compass, scale bar, and grid lines to the
choropleth map.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by Planning Subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type = "8star", size = 2) +
  tm_grid(alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authority (URA)\n and Population data from Department of Statistics (DOS)",
             position = c("left", "bottom"))
```

To reset the default style, refer to the code chunk below.

```{r}
tmap_style("white")
```

### 21.4.6 Drawing Small Multiple Choropleth Maps

**Small multiple maps**, also referred to as **facet maps**, are
composed of many maps arranged side-by-side and sometimes stacked
vertically. Small multiple maps enable the visualisation of how spatial
relationships change with respect to another variable, such as time.

In **tmap**, small multiple maps can be plotted in three ways:

- by assigning multiple values to at least one of the aesthetic
  arguments,
- by defining a group-by variable in `tm_facets()`, and
- by creating multiple stand-alone maps with `tmap_arrange()`.

#### 21.4.6.1 By Assigning Multiple Values to at Least One Aesthetic Argument

In this example, small multiple choropleth maps are created by defining
`ncols` in `tm_fill()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill(c("YOUNG", "AGED"),
          style = "equal",
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

In this example, small multiple choropleth maps are created by assigning
multiple values to at least one aesthetic argument.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons(c("DEPENDENCY", "AGED"),
              style = c("equal", "quantile"),
              palette = list("Blues", "Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

#### 21.4.6.2 By Defining a Group-by Variable in `tm_facets()`

In this example, multiple small choropleth maps are created using
`tm_facets()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) +
  tm_facets(by = "REGION_N",
            free.coords = TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"),
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

#### 21.4.6.3 By Creating Multiple Stand-alone Maps with `tmap_arrange()`

In this example, multiple small choropleth maps are created by combining
stand-alone maps with `tmap_arrange()`.

```{r}
youngmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("YOUNG",
              style = "quantile",
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020) +
  tm_polygons("AGED",
              style = "quantile",
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp = 1, ncol = 2)
```

### 21.4.7 Mapping Spatial Objects Meeting a Selection Criterion

Instead of creating small multiple choropleth maps, you can also use a
selection function to map spatial objects meeting a particular selection
criterion.

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N == "CENTRAL REGION", ]) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45,
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

## 21.5 Reference

### 21.5.1 All About **tmap**

- [tmap: Thematic Maps in
  R](https://www.jstatsoft.org/article/view/v084i06)
- [tmap on
  CRAN](https://cran.r-project.org/web/packages/tmap/index.html)
- [tmap: get
  started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)
- [tmap: changes in version
  2.0](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-changes-v2.html)
- [tmap: creating thematic maps in a flexible way (useR!
  2015)](http://von-tijn.nl/tijn/research/presentations/tmap_user2015.pdf)
- [Exploring and presenting maps with tmap (useR!
  2017)](http://von-tijn.nl/tijn/research/presentations/tmap_user2017.pdf)

### 21.5.2 Geospatial Data Wrangling

- [sf: Simple Features for
  R](https://cran.r-project.org/web/packages/sf/index.html)
- [Simple Features for R: Standardized Support for Spatial Vector
  Data](https://journal.r-project.org/archive/2018/RJ-2018-009/RJ-2018-009.pdf)
- [Reading, Writing and Converting Simple
  Features](https://cran.r-project.org/web/packages/sf/vignettes/sf2.html)

### 21.5.3 Data Wrangling

- [dplyr](https://dplyr.tidyverse.org/)
- [Tidy
  data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
- [tidyr: Easily Tidy Data with `spread()` and `gather()`
  Functions](https://cran.r-project.org/web/packages/tidyr/tidyr.pdf)
