---
title: "23 Analytical Mapping"
format: html
author: "Trinh Minh Tam"
date: "Jun 16, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
editor: 
  markdown: 
    wrap: 72
---

## 23.1 Overview

### 23.1.1 Objectives

In this in-class exercise, you will gain hands-on experience using appropriate R methods to plot analytical maps.

### 23.1.2 Learning outcome

By the end of this in-class exercise, you will be able to use appropriate functions from **tmap** and **tidyverse** to perform the following tasks:

- Import geospatial data in `rds` format into the R environment.
- Create cartographic-quality choropleth maps using appropriate **tmap** functions.
- Create a rate map.
- Create a percentile map.
- Create a box map.

## 23.2 Getting Started

### 23.2.1 Installing and loading packages

Using the steps learned in the previous lesson, install and load the **sf**, **tmap**, and **tidyverse** packages into the R environment.

```{r}
pacman::p_load(tmap, tidyverse, sf)
```

### 23.2.2 Importing data

For the purpose of this hands-on exercise, a prepared dataset called *NGA_wp.rds* will be used. The dataset is a polygon feature data frame providing information on water points in Nigeria at the LGA level. You can find the dataset in the *rds* sub-directory of the hands-on *data* folder.

Using the appropriate function, import *NGA_wp.rds* into the R environment.

```{r}
NGA_wp <- read_rds("../Data/NGA_wp.rds")
```

## 23.3 Basic Choropleth Mapping

### 23.3.1 Visualising distribution of non-functional water points

Plot a choropleth map showing the distribution of non-functional water points by LGA.

```{r}
p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of functional water points by LGAs")
```

```{r}
p2 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "total_wp",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of total water points by LGAs")
```

```{r}
tmap_arrange(p2, p1, nrow = 1)
```

## 23.4 Choropleth Map for Rates

In our readings we have seen the importance of mapping rates rather than counts. The reason is simple: water points are not equally distributed in space. If we do not account for how many water points are located in an area, we end up mapping the total water point size rather than our topic of interest.

### 23.4.1 Deriving proportion of functional and non-functional water points

We will tabulate the proportion of functional water points and the proportion of non-functional water points in each LGA. In the following code chunk, `mutate()` from the **dplyr** package is used to derive two fields, namely *pct_functional* and *pct_nonfunctional*.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional / total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional / total_wp)
```

### 23.4.2 Plotting a map of rate

Plot a choropleth map showing the distribution of the percentage of functional water points by LGA.

```{r}
tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Rate map of functional water points by LGAs")
```

## 23.5 Extreme Value Maps

Extreme value maps are variations of common choropleth maps where the classification is designed to highlight extreme values at the lower and upper ends of the scale, with the goal of identifying outliers. These maps were developed in the spirit of spatialising EDA — that is, adding spatial features to commonly used approaches in non-spatial EDA (Anselin 1994).

### 23.5.1 Percentile Map

The percentile map is a special type of quantile map with six specific categories: 0–1%, 1–10%, 10–50%, 50–90%, 90–99%, and 99–100%. The corresponding break points can be derived using the base R `quantile()` command, passing an explicit vector of cumulative probabilities as `c(0, .01, .1, .5, .9, .99, 1)`. Note that the begin and end points need to be included.

#### 23.5.1.1 Data Preparation

**Step 1:** Exclude records with `NA` using the code chunk below.

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

**Step 2:** Create a customised classification and extract values.

```{r}
percent <- c(0, .01, .1, .5, .9, .99, 1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[, 1], percent)
```

::: callout-important
When variables are extracted from an `sf` data frame, the geometry is extracted as well. For mapping and spatial manipulation, this is the expected behaviour, but many base R functions cannot deal with the geometry. Specifically, `quantile()` will return an error. As a result, `st_set_geometry(NULL)` is used to drop the geometry field.
:::

#### 23.5.1.2 Why write functions?

Writing a function has three big advantages over copy-and-paste:

- You can give a function an evocative name that makes your code easier to understand.
- As requirements change, you only need to update the code in one place, rather than in many.
- You eliminate the chance of making incidental mistakes when copying and pasting (for example, updating a variable name in one place but not in another).

Source: [Chapter 19: Functions](https://r4ds.had.co.nz/functions.html#functions) of *R for Data Science*.

#### 23.5.1.3 Creating the `get.var` function

First, we will write an R function to extract a variable (e.g. *wp_nonfunctional*) as a vector out of an `sf` data frame.

- **Arguments:**
  - `vname`: variable name (as character, in quotes)
  - `df`: name of the sf data frame
- **Returns:**
  - `v`: vector of values (without a column name)

```{r}
get.var <- function(vname, df) {
  v <- df[vname] %>%
    st_set_geometry(NULL)
  v <- unname(v[, 1])
  return(v)
}
```

#### 23.5.1.4 A percentile mapping function

Next, we will write a percentile mapping function using the code chunk below.

```{r}
percentmap <- function(vnam, df, legtitle = NA, mtitle = "Percentile Map") {
  percent <- c(0, .01, .1, .5, .9, .99, 1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_polygons(vnam,
                title = legtitle,
                breaks = bperc,
                palette = "Blues",
                labels = c("< 1%", "1% - 10%", "10% - 50%",
                           "50% - 90%", "90% - 99%", "> 99%")) +
    tm_borders() +
    tm_layout(main.title = mtitle,
              title.position = c("right", "bottom"))
}
```

#### 23.5.1.5 Test driving the percentile mapping function

To run the function, type the code chunk shown below.

```{r}
percentmap("total_wp", NGA_wp)
```

Note that this is just a bare-bones implementation. Additional arguments — such as the title and legend positioning, to name a few — could be passed to customise various features of the map.

### 23.5.2 Box Map

In essence, a box map is an augmented quartile map with an additional lower and upper category. When there are lower outliers, the starting point for the breaks is the minimum value, and the second break is the lower fence. In contrast, when there are no lower outliers, the starting point for the breaks will be the lower fence, and the second break is the minimum value (there will be no observations falling in the interval between the lower fence and the minimum value).

```{r}
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

- Display summary statistics on a choropleth map using the basic principles of a boxplot.
- To create a box map, a custom breaks specification will be used. However, there is a complication: the break points for the box map vary depending on whether lower or upper outliers are present.

#### 23.5.2.1 Creating the `boxbreaks` function

The code chunk below defines an R function that creates break points for a box map.

- **Arguments:**
  - `v`: vector with observations
  - `mult`: multiplier for IQR (default 1.5)
- **Returns:**
  - `bb`: vector with 7 break points, computed using quartiles and fences

```{r}
boxbreaks <- function(v, mult = 1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialise break points vector
  bb <- vector(mode = "numeric", length = 7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) {  # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

#### 23.5.2.2 Creating the `get.var` function

The code chunk below defines an R function to extract a variable as a vector out of an `sf` data frame.

- **Arguments:**
  - `vname`: variable name (as character, in quotes)
  - `df`: name of the sf data frame
- **Returns:**
  - `v`: vector of values (without a column name)

```{r}
get.var <- function(vname, df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[, 1])
  return(v)
}
```

#### 23.5.2.3 Test driving the newly created function

Let's test the newly created function.

```{r}
var <- get.var("wp_nonfunctional", NGA_wp)
boxbreaks(var)
```

#### 23.5.2.4 Box map function

The code chunk below defines an R function to create a box map.

- **Arguments:**
  - `vnam`: variable name (as character, in quotes)
  - `df`: simple features polygon layer
  - `legtitle`: legend title
  - `mtitle`: map title
  - `mult`: multiplier for IQR
- **Returns:**
  - a `tmap` element (plots a map)

```{r}
boxmap <- function(vnam, df,
                   legtitle = NA,
                   mtitle = "Box Map",
                   mult = 1.5) {
  var <- get.var(vnam, df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
    tm_shape(df) +
    tm_fill(vnam, title = legtitle,
            breaks = bb,
            palette = "Blues",
            labels = c("lower outlier",
                       "< 25%",
                       "25% - 50%",
                       "50% - 75%",
                       "> 75%",
                       "upper outlier")) +
    tm_borders() +
    tm_layout(main.title = mtitle,
              title.position = c("left", "top"))
}
```

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```

## 23.6 References

- Anselin, L. (1994). Exploratory Spatial Data Analysis and Geographic Information Systems.
- Wickham, H. & Grolemund, G. *R for Data Science* — [Chapter 19: Functions](https://r4ds.had.co.nz/functions.html#functions).
