---
title: "16 Treemap Visualisation with R"
format: html
author: "Trinh Minh Tam"
date: "May 11, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
---

## 16.1 Overview

In this hands-on exercise, you will gain hands-on experiences on designing treemap using appropriate R packages. The hands-on exercise consists of three main sections:

1.  How to manipulate transaction data into a treemap structure by using selected functions provided in **dplyr** package.
2.  How to plot static treemap by using **treemap** package.
3.  How to design interactive treemap by using **d3treeR** package.

## 16.2 Installing and Launching R Packages

Before we get started, you are required to check if **treemap**, **treemapify** and **tidyverse** packages have been installed in your R.

::: {style="font-size:1.4em"}
```{r}
pacman::p_load(treemap, treemapify, tidyverse)
```
:::

## 16.3 Data Wrangling

In this exercise, *REALIS2018.csv* data will be used. This dataset provides information of private property transaction records in 2018. The dataset is extracted from the REALIS portal of the Urban Redevelopment Authority (URA).

### 16.3.1 Importing the data set

In the code chunk below, `read_csv()` of **readr** is used to import `realis2018.csv` into R and parsed it into tibble R data.frame format.

::: {style="font-size:1.4em"}
```{r}
realis2018 <- read_csv("../Data/realis2018.csv")
```
:::

The output tibble data.frame is called *realis2018*.

### 16.3.2 Data Wrangling and Manipulation

The data.frame *realis2018* is in transaction record form, which is highly disaggregated and not appropriate to be used to plot a treemap. In this section, we will perform the following steps to manipulate and prepare a data.frame that is appropriate for treemap visualisation:

-   Group transaction records by *Project Name*, *Planning Region*, *Planning Area*, *Property Type* and *Type of Sale*, and
-   Compute *Total Unit Sold*, *Total Area*, *Median Unit Price* and *Median Transacted Price* by applying appropriate summary statistics on *No. of Units*, *Area (sqm)*, *Unit Price (\$ psm)* and *Transacted Price (\$)* respectively.

Two key verbs of **dplyr** package, namely `group_by()` and `summarise()`, will be used to perform these steps.

`group_by()` breaks down a data.frame into specified groups of rows. When you then apply the verbs above on the resulting object they'll be automatically applied "by group".

Grouping affects the verbs as follows:

-   Grouped `select()` is the same as ungrouped `select()`, except that grouping variables are always retained.
-   Grouped `arrange()` is the same as ungrouped; unless you set `.by_group = TRUE`, in which case it orders first by the grouping variables.
-   `mutate()` and `filter()` are most useful in conjunction with window functions (like `rank()`, or `min(x) == x`).
-   `sample_n()` and `sample_frac()` sample the specified number/fraction of rows in each group.
-   `summarise()` computes the summary for each group.

In our case, *group_by()* will used together with *summarise()* to derive the summarised data.frame.

::: callout-tip
## Recommendation

Students who are new to **dplyr** methods should consult [Introduction to dplyr](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html) before moving on to the next section.
:::

### 16.3.3 Grouped summaries without the Pipe

The code chunk below shows a typical two-line code approach to perform the steps.

```{r}
realis2018_grouped <- group_by(realis2018, `Project Name`,
                               `Planning Region`, `Planning Area`,
                               `Property Type`, `Type of Sale`)
realis2018_summarised <- summarise(realis2018_grouped,
                          `Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
                          `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
                          `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
                          `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

::: callout-note
-   Aggregation functions such as `sum()` and `median()` obey the usual rule of missing values: if there's any missing value in the input, the output will be a missing value. The argument `na.rm = TRUE` removes the missing values prior to computation.
:::

The code chunk above is not very efficient because we have to give each intermediate data.frame a name, even though we don't have to care about it.

### 16.3.4 Grouped summaries with the *pipe*

The code chunk below shows a more efficient way to tackle the same processes by using the *pipe*, `%>%`:

::: callout-tip
## Recommendation

To learn more about pipe, visit this excellent article: [Pipes in R Tutorial For Beginners](https://www.datacamp.com/community/tutorials/pipe-r-tutorial).
:::

```{r}
realis2018_summarised <- realis2018 %>%
  group_by(`Project Name`, `Planning Region`,
           `Planning Area`, `Property Type`,
           `Type of Sale`) %>%
  summarise(`Total Unit Sold` = sum(`No. of Units`, na.rm = TRUE),
            `Total Area` = sum(`Area (sqm)`, na.rm = TRUE),
            `Median Unit Price ($ psm)` = median(`Unit Price ($ psm)`, na.rm = TRUE),
            `Median Transacted Price` = median(`Transacted Price ($)`, na.rm = TRUE))
```

## 16.4 Designing Treemap with treemap Package

**treemap** package is a R package specially designed to offer great flexibility in drawing treemaps. The core function, *treemap()*, offers at least 43 arguments. In this section, we will only explore the major arguments for designing elegant and yet truthful treemaps.

### 16.4.1 Designing a static treemap

In this section, *treemap()* of **treemap** package is used to plot a treemap showing the distribution of median unit prices and total unit sold of resale condominium by geographic hierarchy in 2017.

First, we will select records belonging to resale condominium property type from *realis2018_summarised* data frame.

```{r}
realis2018_selected <- realis2018_summarised %>%
  filter(`Property Type` == "Condominium", `Type of Sale` == "Resale")
```

### 16.4.2 Using the basic arguments

The code chunk below designs a treemap by using three core arguments of `treemap()`, namely: *index*, *vSize* and *vColor*.

```{r}
treemap(realis2018_selected,
        index  = c("Planning Region", "Planning Area", "Project Name"),
        vSize  = "Total Unit Sold",
        vColor = "Median Unit Price ($ psm)",
        title  = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

::: callout-note
## Things to learn from the three arguments used:

-   **index**: The index vector must consist of at least two column names or else no hierarchy treemap will be plotted. If multiple column names are provided, the first name is the highest aggregation level, the second name the second highest, and so on.
-   **vSize**: The column must not contain negative values. Its values will be used to map the sizes of the rectangles of the treemap.
:::

::: callout-warning
The treemap above was wrongly coloured. For a correctly designed treemap, the colours of the rectangles should be in different intensities showing, in our case, median unit prices.

For `treemap()`, *vColor* is used in combination with the argument *type* to determine the colours of the rectangles. Without defining *type*, `treemap()` assumes `type = "index"`, which maps colour to the hierarchy of planning areas.
:::

### 16.4.3 Working with *vColor* and *type* arguments

In the code chunk below, the *type* argument is defined as `"value"`.

```{r}
treemap(realis2018_selected,
        index  = c("Planning Region", "Planning Area", "Project Name"),
        vSize  = "Total Unit Sold",
        vColor = "Median Unit Price ($ psm)",
        type   = "value",
        title  = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

::: callout-note
## Things to learn from the code chunk above:

-   The rectangles are coloured with different intensities of green, reflecting their respective median unit prices.
-   The legend reveals that the values are binned into ten bins (0–5000, 5000–10000, etc.) with an equal interval of 5000.
:::

### 16.4.4 Colours in treemap package

There are two arguments that determine the mapping to colour palettes: *mapping* and *palette*. The only difference between `"value"` and `"manual"` is the default value for mapping.

-   The `"value"` treemap considers `palette` to be a diverging colour palette (e.g. ColorBrewer's `"RdYlBu"`), and maps it so that 0 corresponds to the middle colour, −max(abs(values)) to the left-end colour, and max(abs(values)) to the right-end colour.
-   The `"manual"` treemap simply maps min(values) to the left-end colour, max(values) to the right-end colour, and mean(range(values)) to the middle colour.

### 16.4.5 The "value" type treemap

The code chunk below shows a value type treemap.

```{r}
treemap(realis2018_selected,
        index   = c("Planning Region", "Planning Area", "Project Name"),
        vSize   = "Total Unit Sold",
        vColor  = "Median Unit Price ($ psm)",
        type    = "value",
        palette = "RdYlBu",
        title   = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

::: callout-note
## Things to learn from the code chunk above:

-   Although the colour palette used is `RdYlBu`, there are no red rectangles in the treemap because all the median unit prices are positive.
-   The reason we see only 5000 to 45000 in the legend is because the *range* argument defaults to `c(min(values), max(values))` with some pretty rounding.
:::

### 16.4.6 The "manual" type treemap

The `"manual"` type does not interpret the values as the `"value"` type does. Instead, the value range is mapped linearly to the colour palette.

The code chunk below shows a manual type treemap.

```{r}
treemap(realis2018_selected,
        index   = c("Planning Region", "Planning Area", "Project Name"),
        vSize   = "Total Unit Sold",
        vColor  = "Median Unit Price ($ psm)",
        type    = "manual",
        palette = "RdYlBu",
        title   = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

::: callout-note
## Things to learn from the code chunk above:

-   The colour scheme used is very confusing. This is because `mapping = (min(values), mean(range(values)), max(values))`. It is not wise to use a diverging colour palette such as `RdYlBu` if the values are all positive or negative.
:::

To overcome this problem, a single colour palette such as `Blues` should be used.

```{r}
treemap(realis2018_selected,
        index   = c("Planning Region", "Planning Area", "Project Name"),
        vSize   = "Total Unit Sold",
        vColor  = "Median Unit Price ($ psm)",
        type    = "manual",
        palette = "Blues",
        title   = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 16.4.7 Treemap Layout

`treemap()` supports two popular treemap layouts: `"squarified"` and `"pivotSize"`. The default is `"pivotSize"`.

-   The **squarified** treemap algorithm (Bruls et al., 2000) produces good aspect ratios, but ignores the sorting order of the rectangles (sortID).
-   The **pivotSize** (ordered treemap) algorithm (Bederson et al., 2002) takes the sorting order (sortID) into account while aspect ratios are still acceptable.

### 16.4.8 Working with *algorithm* argument

The code chunk below plots a squarified treemap by changing the *algorithm* argument.

```{r}
treemap(realis2018_selected,
        index     = c("Planning Region", "Planning Area", "Project Name"),
        vSize     = "Total Unit Sold",
        vColor    = "Median Unit Price ($ psm)",
        type      = "manual",
        palette   = "Blues",
        algorithm = "squarified",
        title     = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

### 16.4.9 Using *sortID*

When the `"pivotSize"` algorithm is used, the *sortID* argument can be used to determine the order in which the rectangles are placed from top left to bottom right.

```{r}
treemap(realis2018_selected,
        index     = c("Planning Region", "Planning Area", "Project Name"),
        vSize     = "Total Unit Sold",
        vColor    = "Median Unit Price ($ psm)",
        type      = "manual",
        palette   = "Blues",
        algorithm = "pivotSize",
        sortID    = "Median Transacted Price",
        title     = "Resale Condominium by Planning Region and Area, 2017",
        title.legend = "Median Unit Price (S$ per sq. m)"
        )
```

## 16.5 Designing Treemap using **treemapify** Package

**treemapify** is a R package specially developed to draw treemaps in **ggplot2**. In this section, you will learn how to design treemaps closely resembling those in the previous section by using **treemapify**. Before you get started, you should read [Introduction to "treemapify"](https://cran.r-project.org/web/packages/treemapify/vignettes/introduction-to-treemapify.html) and its [user guide](https://cran.r-project.org/web/packages/treemapify/treemapify.pdf).

### 16.5.1 Designing a basic treemap

```{r}
ggplot(data = realis2018_selected,
       aes(area = `Total Unit Sold`,
           fill = `Median Unit Price ($ psm)`),
       layout = "scol",
       start  = "bottomleft") +
  geom_treemap() +
  scale_fill_gradient(low = "light blue", high = "blue")
```

### 16.5.2 Defining hierarchy

**Group by Planning Region**

```{r}
ggplot(data = realis2018_selected,
       aes(area     = `Total Unit Sold`,
           fill     = `Median Unit Price ($ psm)`,
           subgroup = `Planning Region`),
       start = "topleft") +
  geom_treemap()
```

**Group by Planning Area**

```{r}
ggplot(data = realis2018_selected,
       aes(area      = `Total Unit Sold`,
           fill      = `Median Unit Price ($ psm)`,
           subgroup  = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap()
```

**Adding boundary lines**

```{r}
ggplot(data = realis2018_selected,
       aes(area      = `Total Unit Sold`,
           fill      = `Median Unit Price ($ psm)`,
           subgroup  = `Planning Region`,
           subgroup2 = `Planning Area`)) +
  geom_treemap() +
  geom_treemap_subgroup2_border(colour = "gray40",
                                size   = 2) +
  geom_treemap_subgroup_border(colour = "gray20")
```

## 16.6 Designing Interactive Treemap using **d3treeR**

### 16.6.1 Installing *d3treeR* package

**d3treeR** is not available on CRAN and must be installed from GitHub.

1.  If this is the first time you install a package from GitHub, install *devtools* first:

```{r}
#| eval: false
install.packages("devtools")
```

2.  Load *devtools* and install **d3treeR** from GitHub:

```{r}
library(devtools)
install_github("timelyportfolio/d3treeR")
```

3.  Launch *d3treeR*:

```{r}
library(d3treeR)
```

### 16.6.2 Designing An Interactive Treemap

The code below performs two processes.

**Step 1:** Use `treemap()` to build a treemap from *realis2018_summarised* and save it as an object called *tm*.

```{r}
tm <- treemap(realis2018_summarised,
              index  = c("Planning Region", "Planning Area"),
              vSize  = "Total Unit Sold",
              vColor = "Median Unit Price ($ psm)",
              type   = "value",
              title  = "Private Residential Property Sold, 2017",
              title.legend = "Median Unit Price (S$ per sq. m)"
              )
```

**Step 2:** Use `d3tree()` to build an interactive treemap from the *tm* object.

```{r}
d3tree(tm, rootname = "Singapore")
```
